Relational programming in Smalltalk

### 2018-12-12 booklet/article.tex
Add more comments on references.

I believe \textit{microkanren} is, first of all, an \textit{educational beast},
concerning unification, lazy streams, backtracking and optimization; the
abstract definition was shown in 
I repeat the exercise of writing it in \citep{Nocentini:kanrens}:
\begin{itemize}
\item Python, native \verb|generator|s :) limits on recursive calls :(
\item OCaml, algebraic datatypes :) hard to extend :(
\item Smalltalk, simple, fast and clear :) many dispatching msgs :/
\end{itemize}

\begin{example}
In math a relation $P$ is usually characterized by
\begin{displaymath}
\forall a,b,c.\,P(a,b,c) \leftrightarrow a + b = c \quad\text{entails}\quad P(1,2,3)
\end{displaymath}
can be expressed using either the \textit{imperative style}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
a := 1.  b := 2.  c := a + b.  Object assert: [ c = 3 ]
\end{minted}
or the \textit{functional style}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Object assert: [ ([ :a :b | a + b ] value: 1 value: 2) = 3 ]
\end{minted}
or, finally, the \textit{declarative style}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Object assert: [ [ :a :b :c | a + b = c ] value: 1 value: 2 value: 3 ]
\end{minted}
which closely resembles the mathematical statement.
\end{example}

>>>>

Another implementation using the \emph{Clojure} dialect of Lisp is
\citep{clojure:core.logic} which targets the JVM an provides constraints and
nominal logic programming; on the other hand, \citep{sullivan:microkanrenhs}
is a clean and concise implementation using Haskell, while many others simpler
versions can be found, written for the sake of understanding and, ultimately,
for fun.

Finally, your humble authors believe that $\mu$Kanren is a sophisticated
\textit{educational machine} because it concerns and mix together unification,
lazy streams, backtracking and optimization techniques; moreover, it allows
programmers to write their own version according to their personal taste and
style, ranging from implementations pretty imperative to ones built over
categories.  We repeat the exercise of writing it in \citep{Nocentini:kanrens}
targeting different languages, in particular
\begin{itemize}
\item using \emph{Python} we take advantage of native \verb|generator|s objects
and we provide a \emph{fair} enumeration strategy that the canonical one
hasn't; on the contrary, we experience computational limits because of
\texttt{RecursionError} exceptions. This implementation is complete and
parallels the one shown in the reference book, accompanied by a unit test suite
that covers \emph{all} questions explained in the book and some new
applications such as (i)~finding fixed points of a set of inference rules and
enumeration of classes of combinatorial objects.
\item using \emph{OCaml} the syntax of the language can be cleanly defined by
algebraic datatypes and the type system helps in the design of the required
operators.  While object-oriented languages make extensibility a relative
simple task, this implementation is harder unless we can make changes to basic
types, at least for what concerns the central \emph{unification} procedure.
All of this is preparatory to modify the tactics mechanism of the \emph{HOL
Light} theorem prover in order to introduce the relational paradigm in that
environment.
\end{itemize}
The current work describe a Smalltalk implementation of the basic system
with the aim to learn and experiment with a pure object environment, such as 
Pharo and Squeak. Starting small, the next example we show three different
kinds to perform an identity among \verb|SmallInteger| objects:

\begin{example}
In math a relation $P$ is usually characterized by a statement such as
\begin{displaymath}
\forall a,b,c.\,P(a,b,c) \leftrightarrow a + b = c \quad\text{from which}\quad
P(1,2,3) \quad\text{can be proved.}
\end{displaymath}
The same check can be expressed in Smalltalk using either the
\textit{imperative style}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
a := 1.  b := 2.  c := a + b.  Object assert: [ c = 3 ]
\end{minted}
or the \textit{functional style}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Object assert: [ ([ :a :b | a + b ] value: 1 value: 2) = 3 ]
\end{minted}
or, finally, the \textit{declarative style}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Object assert: [ [ :a :b :c | a + b = c ] value: 1 value: 2 value: 3 ]
\end{minted}
which closely resembles the mathematical statement.
\end{example}

\section{Resolution methods}
